---
title: "Reproducible Research- Project 1"
author: "Sylvie Cottrell"
date: "April 16, 2015"
output:
  html_document:
    keep_md: yes
---

## Loading and preprocessing the data

```{r, echo=TRUE}
library(plyr)
library(dplyr)
library(lubridate)
library(lattice)
```
1 - Load the data
```{r,echo=TRUE}
data <- read.csv("activity.csv")
data <- tbl_df(data)
```
2- Process/transform the data
 
Convert the date values from characters to DateTime objects
```{r ,echo=TRUE}
data <- mutate(data, date = ymd(date))
head(data)
```

## What is mean total number of steps taken per day
1- Calculate the total number of steps taken per day
Take a subset of the data, omitting the missing values

```{r  remove_na, echo=TRUE}
 df <- filter(data, !is.na(steps))
```

Create a table with the number of steps each day
```{r , echo=TRUE}
  by_day <- aggregate(df$steps, list(day = df$date), sum)
  by_day
```
2- Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
hist(by_day$x, main=" Histogram of total num of steps per day", xlab="Number of steps")
```


3- Calculate the mean and median of the total number of steps taken per day
Get the mean
```{r, echo=TRUE}
mean(by_day$x)
```
Get the median

```{r echo=TRUE}
median(by_day$x)
```


Result : The average of steps per day is **10,766 steps**, the median is **10,765 steps**.

## What is the average daily activity pattern?
 1-Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
 
 Create a data frame with the average number of steps per interval
```{r echo=TRUE}
by_interval <- aggregate(df$steps, list(interval = df$interval), mean)
```
 
 round the number of steps and create a data frame take the rounded values of steps
```{r echo=TRUE}
  by_interval <- by_interval %>% 
                 mutate(steps_average = round(x)) %>%
                   select( interval, steps_average)
  head(by_interval)
```
 
 Create the plot
```{r echo=TRUE}
plot(by_interval$interval, by_interval$steps_average, type="l", xlab= "Interval Number", ylab= "Average of Steps taken", col="blue" , lwd=2, main="Times Series Plot - Interval Num vs Average of Steps")
```

 3- Find the maximum value
```{r echo=TRUE}
max(by_interval$steps_average)
```

Find the corresponding interval number
```{r echo=TRUE}
result <- filter(by_interval, steps_average == 206)
result
```

The interval number with the highest number of steps is **interval number 835** with **206 steps**

##Imputing missing values

  1- Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
  
```{r echo=TRUE}
summary(data)
```
 We see that there are **2,304** missing values
 
 2- Devise a strategy for filling in all of the missing values in the dataset.
 
 I will replace the missing values by the average value for the interval number it corresponds to
 
```{r echo=TRUE}
 interval <- aggregate(steps ~ interval, df, mean)
 interval <- interval %>% 
               mutate(avg_steps = round(steps)) %>%
                select(interval, avg_steps)
head(interval)
```

Add the average of steps per interval to the original data frame
```{r echo=TRUE}
 new_df <- merge(data, interval, by = 'interval', all.y = F)
 head(new_df)
```

Replace missing values by the average of the value in the interval

```{r echo=TRUE}
complete_df <- mutate(new_df, steps = ifelse(is.na(steps), avg_steps, steps))
head(complete_df)
```

3- Histogram
```{r echo=TRUE}
 by_day2 <- aggregate(complete_df$steps, list(day = complete_df$date), sum)
 hist(by_day2$x, main="Histogram 2 - with missing values replaced", xlab="Number of steps")
```

Find mean
```{r echo=TRUE}
mean(by_day2$x)
```

Find median

```{r echo=TRUE}
median(by_day2$x)
```
The mean is about the same as before with a value of **10,765.64 **, the median has gone down by 3, from 10,765 to 10,762. 
We see that the mean and median have about the same values with or wihout the missing values

## Are there differences in activity patterns between weekdays and weekends

 1- Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r echo=TRUE}
complete_df <- mutate(complete_df, day_type = ifelse(weekdays(date)== "Saturday" | weekdays(date) == "Sunday", "weekend", "weekday"))
head(complete_df)
```

2- Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

First I make the value in day_type as factors, and create a data frame with average steps per interval

```{r echo=TRUE}
complete_df$day_type <- as.factor(complete_df$day_type)
str(complete_df)
by_interval <-  aggregate(complete_df$steps, list(interval = complete_df$interval, day_type = complete_df$day_typ), mean)
by_interval<- mutate(by_interval, steps_average = round(x))
by_interval <- select(by_interval, interval, steps_average, day_type)
head(by_interval)
```

Create a lattice plot

```{r echo=TRUE}
xyplot( steps_average ~ interval |day_type, data=by_interval,type="l",xlab="5-minute interval",ylab="Average number of steps taken",main="Average daily activity weekends vs weekdays",layout=c(1,2))
```
 
 We see a difference between the average of steps during weekdays and weekends. 
 The average of steps taken is generally higher on weekends, but with a higher peek during the week. 